New Bayes covariant coupling methods to cope with change of priors in one-vs-one classification

### 2025-07-29 coupling.tex
Fixes from proofread (Ondrej)

For instance, a popular alternative is divergence arising from the quadratic  score \cite{gneiting2007strictly} (also known as Brier score \cite{brier1950verification}). This divergence takes the form
$$
\divm\nolimits_\textrm{quad} (\boldsymbol{p}, \boldsymbol{q})= \sum (p_i - q_i)^2.
$$

@@

For $K=3$, there are three different arboreal coupling methods, which we shall call $\boldsymbol{s}_1$, $\boldsymbol{s}_2$, $\boldsymbol{s}_3$, where the system \eqref{eq:arb1} represents the coupling method $\boldsymbol{s}_2$.

@@

More generally, to obtain a regular system of equations for  any of $K$ , it is necessary and sufficient that the tree formed by edges for which we include equations from Bradley-Terry system is  a spanning tree. Note that any arboreal method is non-canonical, because there is no natural way to choose a spanning tree in a complete graph. In fact, a stronger statement is true -- no arboreal method is symmetric.
%However, we will see in Section \ref{sec:bc2} that by ``averaging'' several stars we can obtain a canonical

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by $\boldsymbol{s}_i$ the prediction of the arboreal method, where the underlying tree is a star centred at the class $C_i$. The predictions will generally span the vector space $R^K$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $\boldsymbol{s}_i(x)$. Let us now define \emph{radial} coupling method. The  method requires that for the  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\begin{split}
	\hat {\boldsymbol{p}} &= \hat p_1 \boldsymbol{s}_1(x) + \ldots + \hat p_K \boldsymbol{s}_K(x)\\
	1 &= \hat p_1 + \ldots + \hat p_K
	\end{split}
	 \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component of $\hat{\boldsymbol{p}}$ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by ${\boldsymbol{s}}_i$.

@@

A very helpful aid to study Bayes covariant methods is property $BC_{ij}$ (the notation stands for \emph{Bayes covariant when restricted to classes} $i$ and $j$). A coupling method $\boldsymbol{v}$ satisfies $BC_{ij}$ if for any change of priors $\pi$ and any  pairwise data $\boldsymbol{R}$ the odds of classes $C_i$ to $C_j$ predicted by $\boldsymbol{v}(\boldsymbol{R}^\pi)$ and $\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi$ are equal. Using earlier notation for odds (see \eqref{eq:alpha}), one has
\begin{align}
	\alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R}^\pi) \bigr) = \alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R})^\pi \bigr)
\end{align}

@@

\begin{itemize}
	\item[a)] a Bayes covariant coupling method satisfies $BC_{ij}$ for any $i\not= j$ (see Proposition \ref{prop:bcprop} in Appendix \ref{app:bc1}),
	%\item[b)] if a coupling method satisfies $BC_{12}, BC_{23}, \ldots, BC_{(K-1),K}$ then it is Bayes covariant (see Proposition \ref{prop:bcprop}),
	\item[b)] if a coupling method satisfies $BC_{ij}$ for all $i\not= j$ then it is Bayes covariant (see Proposition \ref{prop:bcprop} in Appendix \ref{app:bc1}),
	\item[c)] if a coupling method satisfies $BC_{ij}$ and $BC_{jk}$ then it satisfies $BC_{ik}$ (see Lemma \ref{lem:transitivity} in Appendix \ref{app:bc1}).
\end{itemize}

Now we are ready to prove the our first result.

>>>>

For instance, a popular alternative is divergence arising from the quadratic  score \cite{gneiting2007strictly} (also known as Brier score \cite{brier1950verification}). This divergence takes the form
$$
\divm\nolimits_\textrm{quad} (\boldsymbol{d}, \boldsymbol{d'})= \sum (d_i - d'_i)^2.
$$

@@

For $K=3$, there are three different arboreal coupling methods corresponding to three different spanning trees on three vertices. We shall call them $\boldsymbol{s}_1$, $\boldsymbol{s}_2$, $\boldsymbol{s}_3$, where the coupling method $\boldsymbol{s}_i$ includes only the equations that involve class $i$. In particular, \eqref{eq:arb1} represents the coupling method $\boldsymbol{s}_2$.

@@

More generally, to obtain a regular system of equations for  any of $K$ , it is necessary and sufficient that the tree formed by edges for which we include equations from Bradley-Terry system is  a spanning tree. Note that any arboreal method is non-canonical, because there is no natural way to choose a spanning tree in a complete graph. In fact, a stronger statement is true -- no arboreal method is symmetric. 
To see that, consider the action of the transposition of a vertex $v$ of degree one and another vertex $v'$ of degree $>1$ in a spanning tree. The results of coupling involves different number of equations for classes $C_v$ and $C_v'$ which is a contradiction.
%However, we will see in Section \ref{sec:bc2} that by ``averaging'' several stars we can obtain a canonical

@@

In order to convert Hinton's oracle into a true classification method we can use the principle of self-referentiality.  Denote by $\boldsymbol{s}_i$ the prediction of the arboreal method, where the underlying tree is a star centred at the class $C_i$. The predictions will generally span the vector space $R^K$. In that case the prediction $\hat{\boldsymbol{p}}$ of the method  can be expressed as a linear combination of $\boldsymbol{s}_i(x)$. Let us now define \emph{radial} coupling method. The  method requires that for the  weights we could in fact use the components of $\boldsymbol{p}$ i.e.
\begin{equation}
	\begin{split}
	\hat {\boldsymbol{p}} &= \hat p_1 \boldsymbol{s}_1(x) + \ldots + \hat p_K \boldsymbol{s}_K(x)\\
	1 &= \hat p_1 + \ldots + \hat p_K
	\end{split}
	 \label{eq:radial}
\end{equation}
%
The intuition behind \eqref{eq:radial} is that if some component $\hat p_i$ of $\hat{\boldsymbol{p}} =(\hat p_1, \ldots, \hat p_K) $ is close to 1 (and thus the remaining components are close to zero), then the prediction $\hat{\boldsymbol{p}}$ should be close to the prediction of Hinton's oracle which is provided by ${\boldsymbol{s}}_i$.

@@

A very helpful aid to study Bayes covariant methods is property $BC_{ij}$ (the notation stands for \emph{Bayes covariant when restricted to classes} $i$ and $j$). A coupling method $\boldsymbol{v}$ satisfies $BC_{ij}$ if for any change of priors $\pi$ and any  pairwise data $\boldsymbol{R}$ the odds of classes $C_i$ to $C_j$ predicted by $\boldsymbol{v}(\boldsymbol{R}^\pi)$ and $\bigl(\boldsymbol{v}(\boldsymbol{R})\bigr)^\pi$ are equal. Using earlier notation for odds (see \eqref{eq:alpha}), one requires
\begin{align}
	\alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R}^\pi) \bigr) = \alpha_{ij} \bigl(\boldsymbol{v}(\boldsymbol{R})^\pi \bigr)
\end{align}

@@

\begin{itemize}
	\item[a)] a Bayes covariant coupling method satisfies $BC_{ij}$ for any $i\not= j$ (see Proposition \ref{prop:bcprop} in Appendix \ref{app:bc1}),
	%\item[b)] if a coupling method satisfies $BC_{12}, BC_{23}, \ldots, BC_{(K-1),K}$ then it is Bayes covariant (see Proposition \ref{prop:bcprop}),
	\item[b)] conversely, if a coupling method satisfies $BC_{ij}$ for all $i\not= j$ then it is Bayes covariant (see Proposition \ref{prop:bcprop} in Appendix \ref{app:bc1}),
	\item[c)] if a coupling method satisfies $BC_{ij}$ and $BC_{jk}$ then it satisfies $BC_{ik}$ (see Lemma \ref{lem:transitivity} in Appendix \ref{app:bc1}).
\end{itemize}

Now we are ready to prove the our first key result.

### 2025-07-29 intro.tex
Fixes from proofread (Ondrej)

There is a subtle issue with this approach  when handling probabilistic 
classifiers arising in one-vs-one classification framework.   
One-vs-one classification is a two-step framework to solve 
multi-class classification problems.
In the first step it reduces the multiclass problem to a series of binary
classification problems. 
In the second step it deduces the multi-class decision by aggregating decisions 
for the binary problems. Since the procedure is two-step, there is a possibility 
to apply the adjustment of priors either to the probabilistic distributions 
obtained after the first step, or after the second step. It is not clear what 
should be the preferred course of action and this problem motivates our paper.

Our approach to the problem is to evaluate practical performance of different 
methods to aggregate binary predictions. Such methods -- called coupling methods
-- first appeared in work of \cite{refregier1991probabilistic} when making 
multi-class classification using neural networks. 
Many authors followed with alternative proposals 
\cite{price1994pairwise, hastie1998classification,  zahorian1999partitioned,
wu2004probability, vsuch2015new, vsuch2016bayes}.

We formulate the underlying principles of one-vs-one probabilistic classification in Section 2. In Section 3 we present  different classes of coupling methods. 
We pay special attention to methods that commute with change of priors -- Bayes covariant methods, 
since such methods provide an unexpected, and elegant solution to our problem. 
We introduce new coupling methods by forming weighted ensembles. We prove such ensemble of Bayes covariant methods is again Bayes covariant
when a normalizing condition on weights hold (Proposition \ref{prop:ensemble}).  
Our experiments are described in Section 4. In Section 5 we summarize our findings, 
provide practical guidance and outline future research directions.

>>>>

There is a subtle issue with this approach  when handling probabilistic 
classifiers arising in \emph{probabilistic} one-vs-one classification framework.   
One-vs-one classification is a two-step framework to solve 
multi-class classification problems.
In the first step it reduces the multiclass problem to a series of binary
classification problems. 
In the second step it deduces the multi-class decision by aggregating decisions 
for the binary problems. Since the procedure is two-step, there is a possibility 
to apply the adjustment of priors either to the probabilistic distributions 
obtained after the first step, or after the second step. It is not clear what 
should be the preferred course of action and this problem motivates our paper. We note that the dilemma does not appear
in \emph{non-probabilistic} one-vs-one classification e.g. when voting is used in the second step \cite{galar2011overview}.

Our approach to the problem is to evaluate practical performance of different 
methods to aggregate probabilistic binary predictions. Such methods -- called coupling methods
-- first appeared in work of \cite{refregier1991probabilistic} when making 
multi-class classification using neural networks. 
Many authors followed with alternative proposals 
\cite{price1994pairwise, hastie1998classification,  zahorian1999partitioned,
wu2004probability, vsuch2015new, vsuch2016bayes}.

We formulate the underlying principles of one-vs-one probabilistic classification in Section 2. In Section 3 we present  different classes of coupling methods. 
We pay special attention to methods that commute with change of priors -- Bayes covariant methods, 
since such methods provide an  elegant solution to our problem. 
We introduce new coupling methods by forming weighted ensembles. We prove such ensemble of Bayes covariant methods is again Bayes covariant
when a normalizing condition on weights holds (Proposition \ref{prop:ensemble}).  
Our experiments are described in Section 4. In Section 5 we summarize our findings, 
provide practical guidance and outline future research directions.

### 2025-07-29 theory.tex
Fixes from proofread (Ondrej)

	From \eqref{eq:bt1} we can deduce the odds of classes $C_j$ to $C_i$ for the Bayes classifier since
	\begin{align}
	\alpha_{ji}(\boldsymbol{p}^\textrm{multi}) = \frac{p_j^\textrm{multi}(x)}{p_i^\textrm{multi}(x)}
	=  \frac{d_j(x) P_j}{d_i(x) P_i} = \frac{1}{p^\textrm{binary}_{ij}(x)}- 1.
	\end{align}
	Knowing the odds of $C_2$ to $C_1$, $C_3$ to $C_2$ up to $C_{K}$ to $C_{K-1}$ together with the requirement that probabilities sum to one uniquely determines the whole vector $\boldsymbol{p}^\Pi(x)$.
\end{proof}
%\begin{proof}
%Indeed, we have 
%\begin{align*}
%\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
%\end{align*}
%\end{proof}

@@

In practice one does not  know true distributions of samples and hence cannot 
construct Bayes classifiers. However, given a sample from the distribution, 
we can train binary classifiers to provide approximations $r_{ij}(x)$ to 
pairwise Bayes classifiers $p_{ij}^\textrm{multi}(x)$. The goal of probabilistic
one-vs-one classification is to deduce for any $x$ in the feature space $X$ an 
estimate $\hat{\boldsymbol{p}}(x)$ for $\boldsymbol{p}^\textrm{multi}(x)$. 
One does it by finding an approximate solution of the system of equations
%\begin{align}
%
%\end{align}
%
%
%Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
%\begin{equation}
%r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}.
%\end{equation}
%
%
%The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.
%
%
%
%Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

@@

A method to solve Bradley-Terry equations is called a \emph{coupling method}. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries being $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{v}$ proceeds as indicated in Figure \ref{fig:coupling}.

@@

The system of Bradley-Terry equations is usually inconsistent and additional assumptions are needed to construct an estimator. In Sections \ref{sec:des:exact} and \ref{sec:des:inexact} we review various approaches to construct coupling methods.

@@

\emph{Symmetry} We can define two natural actions on permutation group on $K$-elements. Suppose $\sigma$ is a permutation on $K$ elements.

\begin{itemize}
	\item Given  distributions on the classes $\boldsymbol{p}= (p_1, \ldots, p_K)$, we can define
	\begin{align}
		\sigma(\boldsymbol{p})= (p_{\sigma(1)},p_{\sigma(2)}, \ldots, p_{\sigma(K)}).
	\end{align}
	 \item $\sigma$ also acts on the matrix of binary predictions. Namely, the entry of $\sigma(\boldsymbol{R})$ in $i$-th row and $j$-th column is defined to be $r_{\sigma(i), \sigma(j)}$.
	 \end{itemize}
We say that a coupling method $V$ is \emph{symmetric} if 
\begin{align}
		\sigma(\boldsymbol{v}(\boldsymbol{R})) = \boldsymbol{v}(\sigma(\boldsymbol{R}))\quad\textrm{for any permutation $\sigma$}.
\end{align}

@@

Let us make this explicit. Let $\boldsymbol{v}$ be a coupling method. Denote by  $\pi$ the change of priors $\pi:\Pi \rightarrow \Pi'$. If we have a  binary classifier predicting distribution $D \propto (p, p')$ on two classes $C_i$ and $C_j$, then we define new distribution $\pi_2^{ij}(D)$ by requiring
$$
\pi_2^{ij}(D) \propto \biggl(p \frac{P'_i}{P_i},p' \frac{P'_j}{P_j}\biggr).
$$
We denote by $\boldsymbol{R}^\pi$ the $K\times K$ matrix with entries $\pi_2^{ij}(r_{ij})$. The value $\boldsymbol{v}(\boldsymbol{R}^\pi)$ represents the effect of change of priors if we apply \eqref{eq:changePrior} \emph{before} coupling.

@@

\emph{Lack of parameters} is strongly beneficial for situations when the dataset is small and it would be imprudent to reserve a sizeable portion to obtain unbiased estimates of the parameters of the coupling method. This issue gains expediency when the underlying binary classification method requires crossvalidation to select some hyperparameters. For example, this is the case of support vector machines, where often one opts for RBF kernel, which requires choosing two hyper-parameters.

>>>>

	From \eqref{eq:bt1} we can deduce the odds of classes $C_j$ to $C_i$ for the Bayes classifier since
	\begin{align}
	\alpha_{ji}(\boldsymbol{p}^\textrm{multi}) = \frac{p_j^\textrm{multi}(x)}{p_i^\textrm{multi}(x)}
	=  \frac{d_j(x) P_j}{d_i(x) P_i} = \frac{1}{p^\textrm{binary}_{ij}(x)}- 1.
	\end{align}
	Knowing the odds of $C_2$ to $C_1$, $C_3$ to $C_2$ up to $C_{K}$ to $C_{K-1}$ together with the requirement that probabilities sum to one uniquely determines the whole vector $\boldsymbol{p}^\textrm{multi}(x)$.
\end{proof}
%\begin{proof}
%Indeed, we have 
%\begin{align*}
%\frac{1}{r_{ij}(x)} - 1 = \frac{p_j^\Pi}{p_i^\Pi}
%\end{align*}
%\end{proof}

@@

In practice one does not  know true distributions of samples and hence cannot 
construct Bayes classifiers. However, given a sample from the distribution, 
we can train binary classifiers to provide approximations $r_{ij}(x)$ to 
pairwise Bayes classifiers $p_{ij}^\textrm{binary}(x)$. The goal of probabilistic
one-vs-one classification is to deduce for any $x$ in the feature space $X$ an 
estimate $\hat{\boldsymbol{p}}(x)$ for $\boldsymbol{p}^\textrm{multi}(x)$. 
One does it by finding an approximate solution of the system of equations
%\begin{align}
%
%\end{align}
%
%
%Suppose that $K$ classes $C_1, \ldots, C_K$ are distributed according to probability distributions $p_i$ on a  space $X$. Let us denote the $r_{ij}(x)$ the output of Bayes (binary) classifier for $x$ in $X$ yielding the probability $p(x \in C_i \mid x\in C_i \cup C_j )$. Then we have
%\begin{equation}
%r_{ij}(x)= \frac {p_i(x)}{p_i(x) + p_j(x)}.
%\end{equation}
%
%
%The system of equations is called Bradley--Terry equations.  The equations in \eqref{eq:bt1}  can be transformed to a system of linear equations. The theoretical basis for one-vs-one classification framework is provided by the following result.
%
%
%
%Suppose we have estimators $\hat r_{ij}$ for Bayes predictions $r_{ij}$ for all pairs $i\not= j$. A \emph{coupling method} obtains a multi-class probabilistic estimate $\boldsymbol{\hat p}= (\hat p_1, \ldots, \hat p_K)$ that satisfy
\begin{equation}
	{r}_{ij}(x) \approx \frac {\hat p_i(x)}{\hat p_i(x) + \hat p_j(x)}. \label{eq:bt2}
\end{equation}

@@

A method to solve Bradley-Terry equations is called a \emph{coupling method}. Its input can be represented by a matrix $\boldsymbol{R}(x)$ with off-diagonal entries representing predicted probabilities $r_{ij}(x)$. Schematically, classification using coupling method $\boldsymbol{v}$ proceeds as indicated in Figure \ref{fig:coupling}.

@@

The system of Bradley-Terry equations is usually inconsistent and additional assumptions are needed to construct an estimator. In Sections \ref{sec:des:exact} and \ref{sec:des:inexact} we review various desiderata used to construct and compare coupling methods.

@@

\emph{Symmetry} refers to transformations of multi-class predictions when permuting classes. We can define two natural actions on permutation group on $K$-elements. Suppose $\sigma$ is a permutation on $K$ elements.

\begin{itemize}
	\item Given  distributions on the classes $\boldsymbol{p}= (p_1, \ldots, p_K)$, we can define
	\begin{align}
		\sigma(\boldsymbol{p})= (p_{\sigma(1)},p_{\sigma(2)}, \ldots, p_{\sigma(K)}).
	\end{align}
	 \item $\sigma$ also acts on the matrix of binary predictions. Namely, the entry of $\sigma(\boldsymbol{R})$ in $i$-th row and $j$-th column is defined to be $r_{\sigma(i), \sigma(j)}$.
	 \end{itemize}
We say that a coupling method $\boldsymbol{v}$ is \emph{symmetric} if 
\begin{align}
		\sigma(\boldsymbol{v}(\boldsymbol{R})) = \boldsymbol{v}(\sigma(\boldsymbol{R}))\quad\textrm{for any permutation $\sigma$}.
\end{align}

@@

Let us make this explicit. Let $\boldsymbol{v}$ be a coupling method. Denote by  $\pi$ the change of priors $\pi:\Pi \rightarrow \Pi'$. If we have a  binary classifier predicting distribution $D \propto (p, p')$ on two classes $C_i$ and $C_j$, then we define new distribution $\pi_{ij}(D)$ by requiring
$$
\pi_{ij}(D) \propto \biggl(p \frac{P'_i}{P_i},p' \frac{P'_j}{P_j}\biggr).
$$
We denote by $\boldsymbol{R}^\pi$ the $K\times K$ matrix with entries $\pi_{ij}(r_{ij})$. The value $\boldsymbol{v}(\boldsymbol{R}^\pi)$ represents the effect of change of priors if we apply \eqref{eq:changePrior} \emph{before} coupling.

@@

\emph{Lack of parameters} is strongly beneficial for situations when the dataset is small and it would be imprudent to reserve a sizeable portion to obtain unbiased estimates of the parameters of the coupling method. This issue gains expediency when the underlying binary classification method requires crossvalidation to select some hyperparameters. For example, this is the case of support vector machines, where often one opts for RBF kernel that requires choosing two hyper-parameters.
